---
title: "linearalgebra"
description: |
  A short description of the post.
author:
  - name: Maximilian Rohde
    url: https://maximilianrohde.com
date: 07-24-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Resources

- https://rstudio-pubs-static.s3.amazonaws.com/100675_0d981bac13684a5b9f7dc64934573988.html
- https://rstudio-pubs-static.s3.amazonaws.com/99415_2b07b7f8b93d4d4ca79ade8bacc4f89d.html
- https://www.math.uh.edu/~jmorgan/Math6397/day13/LinearAlgebraR-Handout.pdf
- https://bendixcarstensen.com/APC/linalg-notes-BxC.pdf

## Prerequisites
The following material should be understandable to anyone with a basic knowledge of R and linear algebra, and it looking to implement basic linear algebra with R.

## Creating vectors
Vectors can be created with the `c()` function.
```{r}
c(1,3,7)
```

For vectors that are a continuous sequence of numbers, you can use the `:` operator.
```{r}
1:8
```

For vectors of repeated numbers, such as a zero vector, use the `rep()` function.
```{r}
# Create a zero vector of length 10
rep(0, times=10)
```

More complicated sequences can be created with the `seq()` function.
```{r}
# Create a sequence from 1 to 3 with a step size of 0.5
seq(1,3, by = 0.5)
```
```{r}
# Create 10 equal spaced numbers between 1 and 3
seq(1,3, length.out = 10)
```

## Creating matrices
There are two common ways to create matrices in R.

The first method is using the `matrix()` function. You pass the elements of the matrix into `matrix()` and specify the number of rows and columns. Note that R fills in the numbers going down each column. This can be unintuitive.

```{r}
# Put the numbers 1 to 12 in 3 rows and 4 columns
matrix(1:12, nrow=3, ncol=4)
```
```{r}
# Put the numbers 1 to 12 in 2 rows and 6 columns
matrix(1:12, nrow=2, ncol=6)
```

The second method is using `rbind()` and `cbind()`. Use `rbind()` to combine vectors into a matrix by row, and `cbind()` to combine vectors into a matrix by column.

```{r}
x1 <- c(1,2,3,4)
x2 <- c(5,6,7,8)
x3 <- c(9,10,11,12) 
```

```{r}
rbind(x1, x2, x3)
```

```{r}
cbind(x1, x2, x3)
```

To create a $n \times n$ identity matrix, use the `diag(n)` function.
```{r}
diag(5)
```

Once a matrix is created, the `dim()` can be used to obtain the dimensions.
```{r}
m <- cbind(x1, x2, x3)

# See that m has 4 rows and 3 columns
dim(m)
```

If you are only interested in the number of rows and columns, use `nrow()` and `ncol()` respectively.
```{r}
nrow(m)
ncol(m)
```


## Matrix indexing
Once you have a matrix in R, how do you subset parts of the matrix? Let's use this matrix, `m` as an example.

```{r}
m <- matrix(1:12, nrow=3, ncol=4)

print(m)
```

The syntax `m[i]` selects the ith element. Recall that R counts going down columns.
```{r}
m[1]
m[2]
m[3]
m[4]
```

The syntax `m[i, ]` selects the ith row
```{r}
m[1,]
m[2,]
m[3,]
```

The syntax `m[,i]` selects the ith column.
```{r}
m[,1]
m[,2]
m[,3]
m[,4]
```

The syntax `m[i,j]` selects the element in the ith row and jth column.
```{r}
m[1,2]
m[2,3]
m[3,4]
```

## Matrix operations


